### 2015-05-15 intro.tex

Concurrency bugs are exaccerbated by the complex and hostile environment in which drivers operate~\cite{corbet2005linux}. The OS can invoke drivers from multiple threads, a hardware device can issue interrupt requests that cause running processes to block and switch execution context, and the user may remove a device (hot-plugging) while some operation is still running.  These scenarios can cause \emph{data races} if insufficient synchronization mechanisms are in place to control concurrent access to shared resources.
%
Data races are a source of undefined behavior in C, and lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and fix, especially in the context of complex operating systems.
%
Several techniques have been successfully used to analyze device drivers~\cite{ball2006thorough, clarke2004predicate, engler2000checking, henzinger2002temporal, cook2006termination, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral}, but most focus on generic sequential program properties and protocol bugs. Linux kernel analyzers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, can find deadlocks in kernel source code, but are unable to detect races. Techniques that can detect races in drivers~\cite{qadeer2004kiss, pratikakis2006locksmith, voung2007relay, lal2012corral} are usually either \emph{unsound} (i.e.\ can miss real bugs) or \emph{imprecise} (i.e.\ can report false bugs), and typically sacrifice precision for scalability. Thus, there is a clear need for new tools that are able to detect races efficiently and precisely.

We present \whoop, an automated approach for static data race analysis in device drivers. \whoop is empowered by \emph{symbolic pairwise lockset analysis}, which attempts to prove a driver race-free by: (i) deriving a sound \emph{sequential} program that \emph{over-approximates} the originally concurrent program; (ii) instrumenting it to record \emph{locksets}; and (iii) using the locksets to assert that all accesses to the same shared resource are consistently protected by a common lock. Reducing analysis to reasoning over a sequential program avoids the need to enumerate thread interleavings, and allows reuse of existing successful sequential verification techniques.
%
We show that we can exploit the race-freedom guarantees provided by our symbolic analysis to achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a state-of-the-art precise bug-finder used by Microsoft to analyze drivers and other concurrent programs. Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution.  By combining \whoop and \corral we achieve analysis speedups in the range of 1.5--10$\times$ for most of our benchmarks, compared with using \corral in isolation.  For two drivers, we observed even greater speedups of 12$\times$ and 20$\times$.
%
\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems as well as to concurrent systems that use a similar programming model (e.g.\ filesystems).

>>>>

Concurrency bugs are exacerbated by the complex and hostile environment in which drivers operate~\cite{corbet2005linux}. The OS can invoke drivers from multiple threads, a hardware device can issue interrupt requests that cause running processes to block and switch execution context, and the user may remove a device (hot-plugging) while some operation is still running.  These scenarios can cause \emph{data races} if insufficient synchronization mechanisms are in place to control concurrent access to shared resources.
%
Data races are a source of undefined behavior in C~\cite[p.\ 38]{iso/iec11}, and lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and fix, especially in the context of complex operating systems.
%
Several techniques have been successfully used to analyze device drivers~\cite{ball2006thorough, clarke2004predicate, engler2000checking, henzinger2002temporal, cook2006termination, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral}, but most focus on generic sequential program properties and protocol bugs. Linux kernel analyzers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, can find deadlocks in kernel source code, but are unable to detect races. Techniques that can detect races in drivers~\cite{qadeer2004kiss, pratikakis2006locksmith, voung2007relay, lal2012corral} are usually either \emph{unsound} (i.e.\ can miss real bugs) or \emph{imprecise} (i.e.\ can report false bugs), and typically sacrifice precision for scalability. Thus, there is a clear need for new tools that are able to detect races efficiently and precisely.

We present \whoop, an automated approach for static data race analysis in device drivers. \whoop is empowered by \emph{symbolic pairwise lockset analysis}, which attempts to prove a driver race-free by: (i) deriving a sound \emph{sequential} program that \emph{over-approximates} the originally concurrent program; (ii) instrumenting it to record \emph{locksets}; and (iii) using the locksets to assert that all accesses to the same shared resource are consistently protected by a common lock. Reducing analysis to reasoning over a sequential program avoids the need to enumerate thread interleavings, and allows reuse of existing successful sequential verification techniques.
%
We show that we can exploit the race-freedom guarantees provided by our symbolic analysis to achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a state-of-the-art precise bug-finder used by Microsoft to analyze drivers and other concurrent programs. Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution.  By combining \whoop and \corral we achieve analysis speedups in the range of 1.5--10$\times$ for most of our benchmarks, compared with using \corral in isolation.  For two drivers, we observed even greater speedups of 12$\times$ and 20$\times$.
%
\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems as well as to concurrent systems that use a similar programming model (e.g.\ file systems).

### 2015-05-15 whoop.tex
Removed sentence

\noindent\textbf{Sequentialisation }
%
The pseudocode of Figure~\ref{fig:sequentialization} shows the sequential program that we analyze in order to prove race-freedom for a pair of thread templates $T$ and $U$. Assuming that $T$ and $U$ have been instrumented using distinct symbols $i$ and $j$, yielding $\instrument{T}{i}$ and $\instrument{U}{j}$, the sequential program operates as follows. First, the read, write and current locksets for $\instrument{T}{i}$ and $\instrument{U}{j}$ are initialized to be empty, and for each shared variable $s$, the locksets $\lockset{s}{i}$ and $\lockset{s}{j}$ are initialized to the full set of mutexes, $\mutexes$.  The main procedures of the instrumented thread templates, $\mainforthread{\instrument{T}{i}}$ and $\mainforthread{\instrument{U}{j}}$, are then executed in turn (the order does not matter, due to renaming). Finally, an assertion checks for consistent use of mutexes: if $s$ is written during execution of $\instrument{T}{i}$ and accessed during execution of $\instrument{U}{j}$, or vice-versa, then the locksets $\lockset{s}{i}$ and $\lockset{s}{j}$ must contain at least one common mutex.

@@

Houdini does not generate the initial pool of candidates: \whoop generates them using a set of heuristics, and passes them to Houdini as a starting point.  The idea is to automatically generate likely program invariants based on syntactic patterns extracted from an inter-procedural pass over the code for an entry point.  We give two examples; for clarity we use notation from the simple shared variable concurrent programming model of \S\ref{sec:symbolicpairwise}.  If we observe syntactically that procedure $f$ of entry point $T$ may write to, but does not read from, shared variable $s$, then when instrumenting $T$ with symbol $i$, we guess $s \in \hasbeenwritten{i}$ and $s \notin\hasbeenread{i}$ as postconditions for $\instrument{f}{i}$.  These guesses may be incorrect, for instance if the potential write to $s$ turns out to be in dead code, or if a read from $s$ has already been issued on entry to $\instrument{f}{i}$.  Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a postcondition for $\instrument{f}{i}$; this guess may be wrong, e.g.\ if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.  We stress that guessing incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculatively generate candidates that are later deemed to be incorrect, and thus discarded, by Houdini.  The balance we try to strike is to have \whoop generate sufficient candidates to enable precise lockset analysis, without generating so many candidates that the speed of the Houdini algorithm is prohibitively slow.

@@

\noindent\textbf{Limitations }
%
As a lockset analyzer, \whoop can be imprecise because a violation of the locking discipline does not always correspond to a real race (e.g.\ when lockfree synchronization is used). \whoop also uses over-approximation, which can be another source of imprecision. Furthermore, the tool does not check for dynamically created locks or for locks provided by external libraries, although the later could be addressed by providing a mechanism for users to declare custom locks. We also do not currently treat interrupt handlers in a special way; we just assume that they can execute concurrently with any entry point. One way to address this is to model interrupt-specific kernel functions (e.g. for enabling/disabling interrupts).

Another limitation of \whoop is that it is unable to verify drivers designed to be accessed by only a single process at a time. This \emph{single-open device}~\cite{corbet2005linux} mode can be enforced by atomically testing (at the beginning of an entry point) a flag that indicates device availability: if the flag is set to true, then the checking entry point executes, else it blocks. Because \whoop performs pair-wise analysis, it over-approximates this flag, and thus can falsely report a pair as racy. However,~\cite{corbet2005linux} advises against serializing drivers in this way, as it hinders user ingenuity (e.g.\ a user might expect that a device can be accessed concurrently for performance). Because this serialization is considered bad practice, we burden the developer with disregarding any related false bug reports.

>>>>

\noindent\textbf{Sequentialization }
%
The pseudocode of Figure~\ref{fig:sequentialization} shows the sequential program that we analyze in order to prove race-freedom for a pair of thread templates $T$ and $U$. Assuming that $T$ and $U$ have been instrumented using distinct symbols $i$ and $j$, yielding $\instrument{T}{i}$ and $\instrument{U}{j}$, the sequential program operates as follows. First, the read, write and current locksets for $\instrument{T}{i}$ and $\instrument{U}{j}$ are initialized to be empty, and for each shared variable $s$, the locksets $\lockset{s}{i}$ and $\lockset{s}{j}$ are initialized to the full set of mutexes, $\mutexes$.  The main procedures of the instrumented thread templates, $\mainforthread{\instrument{T}{i}}$ and $\mainforthread{\instrument{U}{j}}$, are then executed in turn (the order does not matter, due to renaming). Finally, an assertion checks for consistent use of mutexes: if $s$ is written during execution of $\instrument{T}{i}$ and accessed during execution of $\instrument{U}{j}$, or vice-versa, then the locksets $\lockset{s}{i}$ and $\lockset{s}{j}$ must contain at least one common mutex.

@@

Houdini does not generate the initial pool of candidates: \whoop generates them using a set of heuristics, and passes them to Houdini as a starting point.  The idea is to automatically generate likely program invariants based on syntactic patterns extracted from an inter-procedural pass over the code for an entry point.  We give two examples; for clarity we use notation from the simple shared variable concurrent programming model of \S\ref{sec:symbolicpairwise}.  If we observe syntactically that procedure $f$ of entry point $T$ may write to, but does not read from, shared variable $s$, then when instrumenting $T$ with symbol $i$, we guess $s \in \hasbeenwritten{i}$ and $s \notin\hasbeenread{i}$ as post-conditions for $\instrument{f}{i}$.  These guesses may be incorrect, for instance if the potential write to $s$ turns out to be in dead code, or if a read from $s$ has already been issued on entry to $\instrument{f}{i}$.  Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a post-condition for $\instrument{f}{i}$; this guess may be wrong, e.g.\ if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.  We stress that guessing incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculatively generate candidates that are later deemed to be incorrect, and thus discarded, by Houdini.  The balance we try to strike is to have \whoop generate sufficient candidates to enable precise lockset analysis, without generating so many candidates that the speed of the Houdini algorithm is prohibitively slow.

@@

\noindent\textbf{Limitations }
%
As a lockset analyzer, \whoop can be imprecise because a violation of the locking discipline does not always correspond to a real race (e.g.\ when lock-free synchronization is used). \whoop also uses over-approximation, which can be another source of imprecision. Furthermore, the tool does not check for dynamically created locks or for locks provided by external libraries, although the later could be addressed by providing a mechanism for users to declare custom locks. We also do not currently treat interrupt handlers in a special way; we just assume that they can execute concurrently with any entry point. One way to address this is to model interrupt-specific kernel functions (e.g. for enabling/disabling interrupts).

Another limitation of \whoop is that it is unable to verify drivers designed to be accessed by only a single process at a time. This \emph{single-open device}~\cite{corbet2005linux} mode can be enforced by atomically testing (at the beginning of an entry point) a flag that indicates device availability: if the flag is set to true, then the checking entry point executes, else it blocks. Because \whoop performs pair-wise analysis, it over-approximates this flag, and thus can falsely report a pair as racy. However,~\cite{corbet2005linux} advises against serializing drivers in this way, as it hinders user ingenuity (e.g.\ a user might expect that a device can be accessed concurrently for performance).
